Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course

### 2024-08-06 paper/paper.qmd
found sources and started fixing up written bits

---
title: "Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course"
author: 
  - Rohan Alexander
  - Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/llms-achievement. We thank Nathalie Moon for developing and sharing her survey questions. We thank Tiffany Timbers and attendees at JSM 2024 for helpful suggestions. This research is currently under review by the University of Toronto's University Research Ethics Board. Comments can be sent to: rohan.alexander@utoronto.ca."
date: today
date-format: long
abstract: "To help understand the effect of Large Language Models (LLMs) on data science practice we examine the extent to which LLM usage is correlated with the mark that a student gets on a final paper in a classroom data science setting. We find some very mild evidence from this observational study that LLM usage may be associated with better scores, but our main conclusion is that there is no clear relationship between more extensive LLM usage and the student's mark. Despite the classroom setting used for evaluation, the particular activity of interest reflects the work done by professional data scientists. Our finding suggests the need for more extensive work evaluating how LLMs can be integrated into the data science workflow in a way that provides value."
format: pdf
number-sections: true
table-of-contents: false
bibliography: references.bib
---

@@

library(knitr)
library(quanteda)
library(readxl)
library(reshape2)
library(tidytext)
library(tidyverse)

@@

There are many aspects of trustworthy data science, including education, culture, and workflow to name a few. One aspect of special importance is the tools that we use to do data science: the computers, programming languages, settings, and environments. The wide release of user-friendly Large Language Models (LLMs), especially OpenAI's ChatGPT, is a rare instance of a new tool being made available.

Like all new tools, LLMs have created both excitement and apprehension. ChatGPT's public release on November 30, 2022, brought LLMs into the mainstream conversation. By now many people, especially educators and students, have some experience with LLMs, in both personal and professional contexts.

In the context of teaching statistics and data science LLMs could be useful for many tasks. For instance, there is considerable interest in the potential of chatbots and personalized tutors **[ADD CITE]**. ChatGPT has also been the catalyst for many interesting and important conversations around academic integrity **[ADD CITE]**, the development of critical thinking skills **[ADD CITE]**, and what effective learning looks like **[ADD CITE]**.

@@

The tasks involved in a trustworthy data science workflow can be generally broken down into a number of key competencies [@Adhikari2021Interleaving; @Gibbs2021Building]. The potential for LLMs to positively affect students' academic performance in data science can be clearly inferred from already-demonstrated effects in a number of those competencies on adjacent professional fields. The first is programming, which is done when cleaning, analyzing, visualizing data often using languages like R or Python. The second is writing, which is primarily done when communicating results.

In terms of general programming, @Peng_2023 found a positive impact of GitHub Copilot (an LLM-powered programming assistant) on productivity. Specifically, in an experiment involving 95 freelance programmers, they found that that programmers with access to GitHub Copilot completed a standardized programming task 56% faster than the control group. Programmers with less experience saw the greatest improvements in productivity.

@DellAcqua_2023 focus on management consulting tasks and provide evidence that LLMs can improve writing productivity. In a field experiment involving consultants from Boston Consulting Group they found that the use of OpenAI's GPT-4 led to a 25% increase in delivery speed of business tasks (most involving some writing), as well as a 40% increase in human-rated performance on those tasks. Similar to computer programming, these productivity increases were most pronounced for those with below average performance, with their output increasing by 43%.

@@

@Ellis_Slade_2023 take a more optimistic perspective and argue that LLMs are just another technology that will impact statistics and data science education. Similarly, @Tu_2024 acknowledge that LLMs can streamline many parts of a data science workflow. With that in mind, they suggest that data-scientists-in-training should shift their self-perspectives from primarily being an analyst to primarily being a manager responsible for strategic oversight of the analysis.

@@

To understand the current state of LLMs as a tool for trustworthy data science, this paper focuses on the association between student academic performance and their LLM usage. Specifically, we examine the relationship between students' grades and self-reported measures of student LLM usage, as well as student attitudes toward LLMs in general. This is based on final papers and a survey, conducted in a third-year undergraduate data science course at the University of Toronto. By examining how students interact with and perceive LLMs as tools, and how these variables translate into student outcomes, the effects of LLM integration in data science can be more precisely determined and recommendations made for future development.

@@

1. an optional survey; 
2. self-reported LLM usage; and 
3. student marks on the final paper.

@@

By the time they are working on their final paper, students have submitted and received feedback on two previous papers with similar requirements and rubrics to that of the final paper. Each paper has the same basic structure and expectations. Before the final paper is due students have received feedback on all their previous work in the class (including their past papers) and there is an optional two-day period of peer review.

By virtue of pre-requisites of this course the typical student is an upper-year undergraduate. Coding and writing are major parts of the course. Students are welcome to use R or Python, but the majority code in R because that is the programming language currently mostly taught in pre-requisite courses. All writing must be in English. The aim of the course for students to create a public portfolio of work they can use to apply for jobs.

@@

All data cleaning and analysis was done using the `R` statistical programming language [@citeR], and especially uses the `tidyverse` [@tidyverse], `janitor` [@janitor], `reshape2` [@reshape], and `readxl` [@readxl] libraries.

@@

There is a wide distribution of self-reported GPAs (@fig-response-time-and-gpa-2). Seven of the 100 respondents included in the analysis dataset did not report their GPA. The majority of responses cluster around a B (3.0/4.0), and the average is 3.06 with a standard deviation of 0.55. One factor that may affect the range is that the course is required for programs in the Statistics, Mathematics and Computer Science Departments. Self-reported GPAs introduce the possibility of reporting bias. For instance, respondents may have provided their cumulative GPA, their most recent term's GPA, or could have misreported it entirely.

Students from a range of years took the course, however the majority of respondents were in their 3rd or 4th year of study (@tbl-year-dist). The course's prerequisite of general statistics, which is a two-course sequence typically completed by students in their second year, would make it difficult to take this course earlier than 3rd year.

@@

all_data |>
  rename(ai_familiarity = how_familiar_are_you_with_using_generative_ai_tools_such_as_open_a_is_chat_gpt_or_equivalents) |>
  mutate(
    ai_familiarity = case_when(
      ai_familiarity == "Limited to some liberal arts, such as history courses" ~ "Somewhat familiar",
      TRUE ~ ai_familiarity
    )
  ) |>
  count(ai_familiarity) |>
  kable(col.names = c("Extent of AI familiarity", "Number"))

all_data |>
  rename(ai_schoolwork_appropriate = to_what_extent_do_you_think_using_generative_ai_tools_such_as_chat_gpt_by_open_ai_or_equivalents_is_ethical_and_appropriate_for_schoolwork) |>
  mutate(
    ai_schoolwork_appropriate = case_when(
      ai_schoolwork_appropriate == "Depends of what the main goal of schoolwork is" ~ "It depends",
      ai_schoolwork_appropriate == "I believe student can only use it if they are allowed to" ~ "It depends",
      ai_schoolwork_appropriate == "a controllable amount" ~ "It depends",
      ai_schoolwork_appropriate == "appropriate depending on the context" ~ "It depends",
      ai_schoolwork_appropriate == "appropriate with rules for usage" ~ "It depends",
      ai_schoolwork_appropriate == "depends" ~ "It depends",
      ai_schoolwork_appropriate == "depends on how it is used" ~ "It depends",
      ai_schoolwork_appropriate == "half and half" ~ "It depends",
      ai_schoolwork_appropriate == "unsure" ~ "It depends",
      TRUE ~ ai_schoolwork_appropriate
    )
  ) |>
  count(ai_schoolwork_appropriate) |>
  kable(col.names = c("Ethical and appropriate for school?", "Number"))
```

@@

There is some very mild evidence from this observational study that LLM usage may be associated with better scores. However, this should mostly be treated as a direction for future research. More broadly, our work has three major take-aways: 1) LLMs have a major role to play in enhancing the trustworthiness of data science. 2) Making general purpose chatbots more opinionated and knowledgeable about data science would considerably help. 3) Alternative interaction approaches, beyond chatbots, could be especially useful.

@@

Along these lines, we have considered LLM usage for code and writing as equivalent, but they should actually have different impacts depending on student backgrounds. Distinguishing between native and second-language English speakers, and then focusing on differential LLM usage, could add a great deal to the analysis.

>>>>

---
title: "Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course"
author: 
  - Rohan Alexander
  - Luca Carnegie
  - Nathalie Moon
thanks: "Code and data are available at: https://github.com/lcarnegie/llms-achievement. We thank Nathalie Moon for developing and sharing her survey questions. We thank Tiffany Timbers and attendees at JSM 2024 for helpful suggestions. This research is currently under review by the University of Toronto's University Research Ethics Board. Comments can be sent to: rohan.alexander@utoronto.ca."
date: today
date-format: long
abstract: "To help understand the effect of Large Language Models (LLMs) on data science practice we examine the extent to which LLM usage is correlated with the mark that a student gets on a final paper in a classroom data science setting. We find some very mild evidence from this observational study that LLM usage may be associated with better scores, but our main conclusion is that there is no clear relationship between more extensive LLM usage and the student's mark. Despite the classroom setting used for evaluation, the particular activity of interest reflects the work done by professional data scientists. Our finding suggests the need for more extensive work evaluating how LLMs can be integrated into the data science workflow in a way that provides value."
format: pdf
number-sections: true
table-of-contents: false
bibliography: references.bib
---

@@

library(knitr)
library(quanteda)
library(readxl)
library(reshape2)
library(tidytext)
library(tidyverse)
library(tinytex)

@@

There are many aspects of trustworthy data science, including education, culture, and workflow to name a few. One aspect of special importance is the tools that we use to do data science: the computers, programming languages, and environments. The wide release of user-friendly Large Language Models (LLMs), particularly OpenAI's ChatGPT, is a rare instance of a new and powerful tool being made available to add to a data scientist's toolbelt.

Like all new tools, however, LLMs have created both excitement and apprehension. ChatGPT's public release on November 30, 2022, brought LLMs into mainstream conversation, quickly being put to use in both industry and academia. By now, many people, especially educators and students, have some experience with LLMs, in both personal and professional contexts.

In the context of teaching statistics and data science, LLMs could be useful for many tasks. For instance, there is considerable interest in the potential of chatbots to act as personalized tutors for students [@Winkler_2018]. ChatGPT has also been the catalyst for many interesting and important conversations around academic integrity [@Okaibedi_2023], the development of critical thinking skills, and what effective learning looks like [@BaidooAnu_2023].

@@

The tasks involved in a trustworthy data science workflow can be generally broken down into a number of key competencies [@Adhikari2021Interleaving; @Gibbs2021Building]. The first is programming, which is done when cleaning, analyzing, visualizing data often using languages like R or Python. The second is writing, which is primarily done when communicating results. The potential for LLMs to positively affect students' academic performance in data science can be clearly inferred from already-demonstrated effects in a number of those competencies on adjacent professional fields.

In terms of computer programming, @Peng_2023 found a positive impact of GitHub Copilot (an LLM-powered programming assistant) on productivity. Specifically, in an experiment involving 95 freelance programmers, they found that that programmers with access to GitHub Copilot completed a standardized programming task 56% faster than the control group. Programmers with less experience saw the greatest improvements in productivity.

@DellAcqua_2023, focusing on management consulting tasks, provide evidence that LLMs can improve writing productivity. In a field experiment involving consultants from Boston Consulting Group they found that the use of OpenAI's GPT-4 led to a 25% increase in delivery speed of business tasks (most involving some writing), as well as a 40% increase in human-rated performance on those tasks. Similar to computer programming, these productivity increases were most pronounced for those with below average performance, with their output increasing by 43%.

@@

@Ellis_Slade_2023 take a more optimistic perspective and argue that LLMs are just another technology that will impact statistics and data science education. Similarly, @Tu_2024 acknowledge that LLMs can streamline many parts of a data science workflow. With that in mind, they suggest that data-scientists-in-training should shift their self-perspectives from primarily being an 'analyst' to primarily being a 'product manager' responsible for strategic oversight of the analysis carried out by LLMs.

@@

To understand the current state of LLMs as a tool for trustworthy data science, this paper focuses on the association between student academic performance and their LLM usage. Specifically, we examine the relationship between students' grades and self-reported measures of student LLM usage, as well as student attitudes toward LLMs in general. This is based on students' final papers and a survey, conducted in a third-year undergraduate data science course at the University of Toronto. By examining how students interact with and perceive LLMs as tools, and how these variables translate into student outcomes, the effects of LLM integration in data science can be more precisely determined, leading to better recommendations for their future development.

@@

1. an optional survey; 
2. self-reported LLM usage; and 
3. student marks on their final paper.

@@

By the time they are working on their final paper, students have submitted and received feedback on two previous papers with similar requirements and rubrics to that of the final paper. Each paper has the same basic structure and expectations. Before the final paper is due, students have received feedback on all their previous work in the class (including their past papers) and there is an optional two-day period of peer review.

By virtue of pre-requisites of this course the typical student is an upper-year undergraduate. Coding and writing are major parts of the course. Students are welcome to use R or Python, but the majority code in R because that is the programming language currently mostly taught in pre-requisite courses. All writing must be in English. The primary aim of the course for students is to create a public portfolio of work they can use to apply for jobs.

@@

All data cleaning and analysis was done using the `R` statistical programming language [@citeR], and especially uses the `tidyverse` [@tidyverse], `janitor` [@janitor], `reshape2` [@reshape], and `readxl` [@readxl] packages.

@@

There is a wide distribution of self-reported GPAs (@fig-response-time-and-gpa-2). Seven of the 100 respondents included in the analysis dataset did not report their GPA. The majority of responses cluster around a B (3.0/4.0), and the average is 3.06 with a standard deviation of 0.55. One factor that may affect the range is that the course is required for programs in the Statistics, Mathematics and Computer Science Departments. Self-reported GPAs also introduce the possibility of reporting bias. For instance, respondents may have provided their cumulative GPA, their most recent term's GPA, or could have misreported it entirely.

Students from a range of years took the course, however the majority of respondents were in their 3rd or 4th year of study (@tbl-year-dist). The course's prerequisite of general statistics, which is a two-course sequence typically completed by students in their second year, would make it difficult to take this course earlier than their 3rd year.

@@

all_data |>
  rename(ai_familiarity = how_familiar_are_you_with_using_generative_ai_tools_such_as_open_a_is_chat_gpt_or_equivalents) |>
  mutate(
    ai_familiarity = case_when(
      ai_familiarity == "Limited to some liberal arts, such as history courses" ~ "Somewhat familiar",
      TRUE ~ ai_familiarity
    )
  ) |>
  count(ai_familiarity) |>
  kable(col.names = c("Extent of AI familiarity", "Students"))

all_data |>
  rename(ai_schoolwork_appropriate = to_what_extent_do_you_think_using_generative_ai_tools_such_as_chat_gpt_by_open_ai_or_equivalents_is_ethical_and_appropriate_for_schoolwork) |>
  mutate(
    ai_schoolwork_appropriate = case_when(
      ai_schoolwork_appropriate == "Depends of what the main goal of schoolwork is" ~ "It depends",
      ai_schoolwork_appropriate == "I believe student can only use it if they are allowed to" ~ "It depends",
      ai_schoolwork_appropriate == "a controllable amount" ~ "It depends",
      ai_schoolwork_appropriate == "appropriate depending on the context" ~ "It depends",
      ai_schoolwork_appropriate == "appropriate with rules for usage" ~ "It depends",
      ai_schoolwork_appropriate == "depends" ~ "It depends",
      ai_schoolwork_appropriate == "depends on how it is used" ~ "It depends",
      ai_schoolwork_appropriate == "half and half" ~ "It depends",
      ai_schoolwork_appropriate == "unsure" ~ "It depends",
      TRUE ~ ai_schoolwork_appropriate
    )
  ) |>
  count(ai_schoolwork_appropriate) |>
  kable(col.names = c("Ethical & Appropriate for school?", "Students"))
```

@@

There is some very mild evidence from this observational study that LLM usage may be associated with better scores. However, this work should mostly be treated as a direction for future research in this area. More broadly, our work has three major take-aways: 1) LLMs have a major role to play in enhancing the trustworthiness of data science. 2) Making general purpose chatbots more opinionated and knowledgeable about data science would considerably help (i.e. fine-tuning them - maybe talk more about the GPT store and the "Telling Stories with Data chatbot") 3) Alternative interaction approaches, beyond a chatbot interface, could be especially useful.

@@

Along these lines, we have considered LLM usage for code and writing as equivalent, but they should actually have different impacts depending on student backgrounds. Distinguishing between native and second-language English speakers, and then focusing on differential LLM usage, could have added a great deal of nuance to the analysis.
